---
title: "Clustering Homework"
author: "Aaron"
date: "July 18, 2016"
output: 
  html_document: 
    highlight: haddock
    theme: cosmo
---

```{r setup, include=FALSE, message=FALSE, warning=F}
knitr::opts_chunk$set(echo = TRUE)
```

### Homework Assignment: 

- Using the mtcars data set
- Create a kmeans object from the first, second, and third columns
- What is the size of each cluster?
- What are the centers of each cluster?
- What is the average disp, wt, and qsec of each cluster?
- Describe each cluster in English. 

```{r message=F, warning=F}

data("mtcars")

library(dplyr)
library(knitr)
library(ggplot2)
library(GGally)

```

First we need to determine how many clusters to create. 

```{r}
set.seed(1234)
wi = 0

for (i in 2:10) {
  z = kmeans(mtcars[,1:3], i)
  wi[i-1] = z$betweenss / z$totss
}
# report the list of within cluster sum of squares
plot(2:10, wi, type='l', xlab="k=2:10", ylab="Within Sum of Squares")

```

The “withinss” or ‘within sum of squares’ value goes aysmptotic at k=**4**. This implies that with the three columns in the `mtcars` data set, the cars can be grouped into 4 groups. 

### Create a kmeans object from the first, second, and third columns

What is the size of each cluster?


```{r}

r = kmeans(mtcars[,1:3], centers =4, nstart=20)

r$size

```

The size of those 4 clusters are 8, 4,9 and 11.

### What are the centers of each cluster?

```{r}

r$centers

```

### What is the average disp, wt, and qsec of each cluster?

For this task, we should derive the cluster labels as an aggregation dimension for the data set. 

```{r}

mtcars %>% 
    mutate(clusters = r$cluster) %>% 
    group_by(clusters) %>% 
    summarise_each(funs(mean), c(disp, wt, qsec)) %>% kable()

```

### Describe each cluster in English. 

Cluster 1: cars that have medium disp, medium mpg and cyl. 

<br>

Cluster 2: cars that have highest disp, lowest mpg. 

<br>

Cluster 3: car that have lowest disp, highest mpg. 

<br>

Cluster 4: cars that have medium to high disp, low mpg and highest cyl. 

<br>


Visually describe all the clusters:

```{r fig.height= 7, fig.width= 9, message=F, warning=F}

mtcars %>% 
    mutate(cluster = as.factor(r$cluster)) %>% 
    select(c(cluster, disp, wt, qsec)) %>% 
    ggpairs(mapping = aes(color = cluster))

```



```{r fig.height= 7, fig.width= 9, message=F, warning=F}

mtcars %>% 
    mutate(cluster = as.factor(r$cluster)) %>% 
    select(c(cluster, disp, mpg, cyl, disp)) %>% 
    ggpairs(mapping = aes(color = cluster))

```


